Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2021-04-26 analysis/paper/paper.Rmd
Usa visual editor (corta líneas cada 80 caracteres)

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence on temperature and precipitation anomalies, and sea ice concentration [e.g. @fogt2020]. Its positive phase is usually described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

@@

Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region. The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007;@silvestri2009;@fogt2012;@rosso2018]. The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].

Some of the variability associated with the zonal asymmetries of the SAM seems to be forced by the tropics. ENSO-like variability affects the Southern Hemisphere extratopics through the Rossby wave trains [@mo1987;@kidson1988;@karoly1989] which project strongly onto the zonal anomalies associated with the SAM in the Pacific sector. Moreover, tropical influences on the SAM have been observed [@fan2007;@fogt2011; @clem2013]. @fan2007 computed SAM indices of the Western and the Eastern Hemispheres separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.

Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and references therein]. It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013]. However, it’s not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.

@@

Our objective is, then, to describe the zonally asymmetric and symmetric components of the SAM variability. We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively. Their vertical structure and coherence, temporal variability and trends are consequently assessed. We then study the spatial patterns described by the variability exclusive to each index focusing on 50\ hPa as representing the stratosphere and 700\ hPa as representing the troposphere. Finally, the relationships of the SAM at 700\ hPa with temperature and precipitation anomalies are investigated.

@@

We used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.

@@

To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) at 700\ hPa. The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the SAM index, Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM) indices as the coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude). The three indices are then normalized by dividing them by the standard deviation of the SAM index at each level. As a result, the magnitudes between indices are comparable. However, only SAM index has unit standard deviation per definition. The explained variance of each pattern is used as an indicator of the degree of zonally symmetry or asymmetry of each monthly field. To quantify the coherence between temporal series corresponding to different indices or the same index at different levels, we computed the temporal correlation between them.

@@

The method assumes linearity in the asymmetric component of the SAM. That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-). @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites. To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700\ hPa and 50\ hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)). In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign. Spatial correlations between composites for each season are high. The method considered in this study seems then a reasonable approximation of the phenomenon.

By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons. The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently. The following seasons were considered – December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON). Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year. Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA). Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.

@@

Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020]. All graphics are made using ggplot2 [@wickham2009]. We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].

@@

We first asses the temporal evolution of A\nobreakdash-SAM and S\nobreakdash-SAM. Figure \@ref(fig:asymsam-timeseries) shows the corresponding time series for 700\ hPa and 50\ hPa and their corresponding density estimates. We selected these two levels as representative of the tropospheric and stratospheric variability respectively. As it is shown below, the variabilities of both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.

@@

From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987). Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags. Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev`\ hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index. (Correlations at lags -5 to 5 are shown in Figure \@ref(fig:A1).)

@@

Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the SAM, A\nobreakdash-SAM  and S\nobreakdash-SAM indices. For the SAM (Figure \@ref(fig:cross-correlation)a), high values below 100\ hPa reflect the vertical (zero-lag) coherency throughout the troposphere. Above 100\ hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability. But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g. grater than 0.4 between tropospheric levels and levels below 30\ hPa). A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour. Stratospheric coherency is stronger for the A\nobreakdash-SAM than the S\nobreakdash-SAM. The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.

<!-- ```{r} -->
<!-- indices_wide %>%  -->
<!--   .[, FitLm(as.numeric(time), asym, sym, se = TRUE), by = .(lev)] %>%  -->
<!--   rm_intercept() %>%  -->
<!--   # .[, term := NULL] %>%  -->
<!--   .[, estimate := estimate/(3600*24*365*10)] %>% -->
<!--   .[, std.error := std.error/(3600*24*365*10)] %>% -->
<!--   # setnames("type", "term") %>%  -->
<!--   ggplot(aes(lev, estimate)) + -->
<!--   geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) + -->
<!--   geom_line() + -->
<!--   scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) + -->
<!--   geom_hline(yintercept = 0) + -->
<!--   scale_x_level() + -->
<!--   scale_y_continuous("Standard deviations per decade") + -->
<!--   coord_flip() + -->
<!--   facet_grid(~ term,labeller = labeller(term = lab_sam)) + -->
<!--   tag_facets()  -->
<!-- ``` -->

@@

The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)). The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50\ hPa and reaches its maximum value at 100\ hPa. The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100\ hPa maximum is much more defined. In Winter and Spring, we detect no statistically significant trend. This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons [e.g. @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.

@@

We then computed the spatial regression of geopotential height anomalies on the A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700\ hPa and 50\ hPa levels (Figure \@ref(fig:2d-regr)). While regression coefficients in Figure \@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple regression using the A\nobreakdash-SAM  and S\nobreakdash-SAM indices at the same time. Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other one.

@@

The amplitude of the first zonal wavenumbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero represents the amplitude of the zonal mean. Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels. However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`). Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean. On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.

@@

To analyse the vertical structure of the geopotential height anomalies associated with the A\nobreakdash-SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50\ hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and for the 700\ hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)b). The geopotential height anomalies associated with the stratospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric A\nobreakdash-SAM. The vertical structure of this signal tilts about 60\degree\ to the West between 100\ hPa and 1\ hPa, suggesting baroclinic processes. The signal in the stratosphere maximises near 10\ hPa despite using the 50\ hPa index for the regression.

The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has significant signals that extend upwards to the uppermost levels considered. In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250\ hPa. The anomalies are larger in the Western hemisphere, where they extend into the stratosphere. In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere. So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM. The structures shown in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level. For any stratospheric (above 100\ hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10\ hPa in Figure \@ref(fig:vertical-regression)a. Conversely, for any tropospheric (below 100\ hPa) index, the result is very similar to Figure \@ref(fig:vertical-regression)b. The patterns mainly change in amplitude (not shown).

@@

To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700\ hPa. As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.

Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700\ hPa with surface land air and sea temperature for each trimester. In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies. The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E respectively. In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO. Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively. Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics. Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM. This signal is not present in the regression pattern with the SAM. Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.

@@

Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South America respectively. South Africa is not shown because no significant signal was detected there.

@@

pp <- GPCC() %>% 
  ReadNetCDF(subset = list(lat = c(-90, 0),
                           time = range(indices$time))) %>% 
  na.omit() %>% 
  normalise_coords() %>%
  .[, days := lubridate::days_in_month(time[1]), by = time] %>% 
  .[, precip := precip/days]

@@

The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation. There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly affected. However, establishing this link would require further research.

>>>>

The Southern Annular Mode (SAM) is the main mode of variability in the Southern
Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and
decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence
on temperature and precipitation anomalies, and sea ice concentration [e.g.
@fogt2020]. Its positive phase is usually described as anomalously low pressures
over Antarctica surrounded by a ring of anomalous high pressures in
middle-to-high latitudes.

@@

Even though these indices are based on zonal averages, their associated
geopotential height spatial anomalies contain noticeable deviations from zonal
symmetry, particularly in the Pacific Ocean region. The zonal asymmetries have
not been widely studied, but previous work suggest that they strongly modulate
the regional impacts of the SAM [@fan2007; @silvestri2009; @fogt2012;
@rosso2018]. The fact that the SAM is not entirely zonally symmetric hinders our
ability to reconstruct its historical variability prior to the availability of
dense observations in the Southern Hemisphere [@jones2009].

Some of the variability associated with the zonal asymmetries of the SAM seems
to be forced by the tropics. ENSO-like variability affects the Southern
Hemisphere extratopics through the Rossby wave trains [@mo1987; @kidson1988;
@karoly1989] which project strongly onto the zonal anomalies associated with the
SAM in the Pacific sector. Moreover, tropical influences on the SAM have been
observed [@fan2007; @fogt2011; @clem2013]. @fan2007 computed SAM indices of the
Western and the Eastern Hemispheres separately and found that they were much
more correlated to each other if the (linear) signal of the ENSO was removed.

Positive trends in the SAM have been documented by various researchers using
different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and
references therein]. It is thought that these trends are driven primarily by
stratospheric ozone depletion and the increase in greenhouse gases, and
understood in the context of zonal mean variables [@marshall2004; @gillett2005;
@arblaster2006; @gillett2013]. However, it's not clear yet how or if the
asymmetric SAM component responds to these forcings, or how its variability
alters the observed trends.

@@

Our objective is, then, to describe the zonally asymmetric and symmetric
components of the SAM variability. We first propose a methodology that provides
for each level, two indices which aim to capture independently the variability
of the symmetric and asymmetric SAM component respectively. Their vertical
structure and coherence, temporal variability and trends are consequently
assessed. We then study the spatial patterns described by the variability
exclusive to each index focusing on 50 hPa as representing the stratosphere and
700 hPa as representing the troposphere. Finally, the relationships of the SAM
at 700 hPa with temperature and precipitation anomalies are investigated.

@@

We used monthly geopotential height at 2.5\degree longitude by
2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels
from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis
to the post-satellite era to avoid any confounding factors arising from the
incorporation of satellite observations.

@@

To separate the zonally symmetric and asymmetric components of the SAM, we
computed the zonal mean and anomalies of the full SAM spatial pattern, as shown
in Figure \@ref(fig:method) at 700 hPa. The full spatial signal
($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric
($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric
($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the SAM index,
Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM)
indices as the coefficients of the regression of each monthly geopotential
height field on the respective patterns (weighting by the cosine of latitude).
The three indices are then normalized by dividing them by the standard deviation
of the SAM index at each level. As a result, the magnitudes between indices are
comparable. However, only SAM index has unit standard deviation per definition.
The explained variance of each pattern is used as an indicator of the degree of
zonally symmetry or asymmetry of each monthly field. To quantify the coherence
between temporal series corresponding to different indices or the same index at
different levels, we computed the temporal correlation between them.

@@

The method assumes linearity in the asymmetric component of the SAM. That means
that zonal asymmetries associated with positive SAM phase (SAM+) are almost
opposite in sign and of the same magnitude to the ones associated with negative
SAM phase (SAM-). @fogt2012's composites (their Figure 4) suggest that this
might not be entirely valid, although much of that apparent non-linearity could
be due to the heterogenous nature of the selected years for constructing the
composites. To test this assumption, we computed seasonal composites of zonal
anomalies of geopotential height for SAM+ and SAM- (defined as months in which
the SAM index is greater than 1 standard deviation and lower than minus 1
standard deviation, respectively) for the period from 1979 to 2018 at the
700 hPa and 50 hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)). In all
seasons and both levels, SAM+ composites are similar to SAM- in structure but
with the opposite sign. Spatial correlations between composites for each season
are high. The method considered in this study seems then a reasonable
approximation of the phenomenon.

By performing the EOF analysis using data for all months we are assuming that
the zonally asymmetric structure of the SAM is the same at all seasons. The
latter was assessed by computing geopotential height zonal anomalies by
projecting the first EOF of each season independently. The following seasons
were considered -- December to February (DJF), March to May (MAM), June to
August (JJA) and September to November (SON). Results are very similar to each
other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial
correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the
stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons
except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison
with the rest of the year. Spatial correlations in the stratosphere are between
-0.24 (DJF with SON) and 0.95 (MAM with JJA). Therefore, the results confirm
that the zonal asymmetric structure of the SAM is very similar throughout most
of the year with the exception of DJF in the stratosphere.

@@

Statistical significance for regression fields were evaluated adjusting p-values
by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to
avoid misleading results from the high number of regressions [@walker1914;
@katz1991].

@@

We performed all analysis in this paper using the R programming language
[@rcoreteam2020], using the data.table package [@dowle2020] and the metR package
[@campitelli2020]. All graphics are made using ggplot2 [@wickham2009]. We
downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and
indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered
using knitr and rmarkdown [@xie2015; @allaire2019].

@@

We first asses the temporal evolution of A\nobreakdash-SAM and
S\nobreakdash-SAM. Figure \@ref(fig:asymsam-timeseries) shows the corresponding
time series for 700 hPa and 50 hPa and their corresponding density estimates. We
selected these two levels as representative of the tropospheric and
stratospheric variability respectively. As it is shown below, the variabilities
of both indices are highly coherent within each atmospheric layer, therefore is
reasonable to take one level as representative of each layer.

@@

From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time
series appear to be correlated. Moreover, looking at the extremes in the
stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM
series (see, for example, the positive events on late 1987). Figure
\@ref(fig:cor-lev) shows these correlations along all levels considered, for
zero and -1 lags. Values of zero-lag correlations between A\nobreakdash-SAM and
S\nobreakdash-SAM are relatively constant throughout the troposphere,
fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and
`r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations
are similarly constant but significantly reduced to around
`r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere,
zero-lag correlations drop to a minimum of
`r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at
`r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then increase again
monotonically with height up to the uppermost level of the reanalysis (although
results near the top of the models are to be interpreted with care). At the same
time, one-month-lag correlations increase with height. Therefore, stratospheric
A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
(Correlations at lags -5 to 5 are shown in Figure \@ref(fig:A1).)

@@

Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across
levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices. For the SAM
(Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the
vertical (zero-lag) coherency throughout the troposphere. Above 100 hPa,
correlation between levels does falls off more rapidly, indicating less coherent
(zero-lag) variability. But correlations between tropospheric and
lower-to-middle stratospheric levels are still relatively high (e.g. grater than
0.4 between tropospheric levels and levels below 30 hPa). A\nobreakdash-SAM and
S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively)
share similar high level of coherency in the troposphere but they differ in
their stratospheric behaviour. Stratospheric coherency is stronger for the
A\nobreakdash-SAM than the S\nobreakdash-SAM. The stratospheric
S\nobreakdash-SAM seems to connect more strongly to the troposphere than the
stratospheric A\nobreakdash-SAM.

<!-- ```{r} -->

<!-- indices_wide %>%  -->

<!--   .[, FitLm(as.numeric(time), asym, sym, se = TRUE), by = .(lev)] %>%  -->

<!--   rm_intercept() %>%  -->

<!--   # .[, term := NULL] %>%  -->

<!--   .[, estimate := estimate/(3600*24*365*10)] %>% -->

<!--   .[, std.error := std.error/(3600*24*365*10)] %>% -->

<!--   # setnames("type", "term") %>%  -->

<!--   ggplot(aes(lev, estimate)) + -->

<!--   geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) + -->

<!--   geom_line() + -->

<!--   scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) + -->

<!--   geom_hline(yintercept = 0) + -->

<!--   scale_x_level() + -->

<!--   scale_y_continuous("Standard deviations per decade") + -->

<!--   coord_flip() + -->

<!--   facet_grid(~ term,labeller = labeller(term = lab_sam)) + -->

<!--   tag_facets()  -->

<!-- ``` -->

@@

The linear trends for each of the indices (SAM, S\nobreakdash-SAM and
A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each
level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
The SAM index presents a statistically positive significant trend (Figure
\@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends
throughout the troposphere up to about 50 hPa and reaches its maximum value at
100 hPa. The seasonal trends (rest of Figure \@ref(fig:trends) column a)
indicate that positive trends are present in Autumn and particularly in Summer,
where the 100 hPa maximum is much more defined. In Winter and Spring, we detect
no statistically significant trend. This is consistent with the results of
previous studies, which find large positive trends in Summer, smaller in Autumn
and no trends in the other seasons [e.g. @fogt2020 and references therein] using
indices of the SAM based on surface or near-surface circulation.

@@

We then computed the spatial regression of geopotential height anomalies on the
A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700 hPa and 50 hPa levels
(Figure \@ref(fig:2d-regr)). While regression coefficients in Figure
\@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients
in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple
regression using the A\nobreakdash-SAM and S\nobreakdash-SAM indices at the same
time. Thus, they are to be interpreted as the patterns associated with each
index, removing the variability (linearly) explained by the other one.

@@

The amplitude of the first zonal wavenumbers at each latitude at 50 hPa and
700 hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero
represents the amplitude of the zonal mean. Zonal wave amplitudes of the spatial
pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column
`r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at
both levels. However, zonal waves are important, particularly South of 50\degree
S, with wavenumber 1 clearly dominating at 50 hPa (Figure
\@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of
waves of similar amplitude at 700 hPa (Figure
\@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`). Figure
\@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the
A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere
(Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while
in the troposphere it is explained by a combination of zonal waves 3 to 1 in
decreasing level of importance (Figure
\@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with
negligible amplitude of the zonal mean. On the other hand, the S\nobreakdash-SAM
it is almost entirely explained by the zonal mean at both levels (Figure
\@ref(fig:wave-amplitude) column c), with little to no contribution from zonal
waves with wavenumbers 1 to 3.

@@

To analyse the vertical structure of the geopotential height anomalies
associated with the A\nobreakdash-SAM index, we show a vertical cross section of
regressions of mean geopotential height between 65\degree S and 40\degree S for
the 50 hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and
for the 700 hPa A\nobreakdash-SAM index (Figure
\@ref(fig:vertical-regression)b). The geopotential height anomalies associated
with the stratospheric A\nobreakdash-SAM (Figure
\@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere,
which underscores the uncoupling between the stratospheric and tropospheric
A\nobreakdash-SAM. The vertical structure of this signal tilts about
60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic
processes. The signal in the stratosphere maximises near 10 hPa despite using
the 50 hPa index for the regression.

The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has
significant signals that extend upwards to the uppermost levels considered. In
the troposphere, the wave-3 structure is equivalent barotropic with maximum
amplitude at roughly 250 hPa. The anomalies are larger in the Western
hemisphere, where they extend into the stratosphere. In the Eastern hemisphere
the wave-3 signal is smaller and confined to the troposphere while negative
anomalies dominate in the stratosphere. So, while the tropospheric
A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies,
these do not project strongly onto the stratospheric A\nobreakdash-SAM. The
structures shown in Figure \@ref(fig:vertical-regression) are robust to the
choice of index level. For any stratospheric (above 100 hPa) index, the
resulting anomalies are very similar to the wave-1 structure with maximum near
10 hPa in Figure \@ref(fig:vertical-regression)a. Conversely, for any
tropospheric (below 100 hPa) index, the result is very similar to Figure
\@ref(fig:vertical-regression)b. The patterns mainly change in amplitude (not
shown).

@@

To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM
and S\nobreakdash-SAM indices we regressed surface land air and sea temperature
and precipitation onto each of the three SAM indices at 700 hPa. As shown in
previous sections, the three indices are highly coherent in the troposphere, so
we select this level to represent the tropospheric circulation for consistency
with previous studies.

Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at
700 hPa with surface land air and sea temperature for each trimester. In Summer,
positive values of the SAM index (Figure
\@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are
associated with negative temperature anomalies near Antarctica which are
surrounded by a ring of positive anomalies. The ring is not zonally symmetric,
as there are four distinctive local maximums around 30\degree W, 120\degree W,
150\degree E and 90\degree E respectively. In the tropics, there are negative
anomalies in the equatorial Pacific, consistent with the negative correlation
between SAM and ENSO. Figure
\@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))`
show temperature anomalies associated with positive values of A\nobreakdash-SAM
and S\nobreakdash-SAM, respectively. Both the local maximums in the ring and the
anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM
regression map, while temperature patterns linked to positive S\nobreakdash-SAM
show a more zonally consistent ring and less relation to the tropics.
Temperature anomalies in the Indian ocean, South Africa and Australia are
strongly related to A\nobreakdash-SAM. This signal is not present in the
regression pattern with the SAM. Spring (Figure \@ref(fig:regr-air-season), row
`r get_row(season == "SON")`) features similar patterns but of smaller
magnitude, with fewer regions where regressed anomalies have statistical
significance.

@@

Regression of the SAM indices with seasonal mean precipitation and 700 hPa
geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and
\@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South
America respectively. South Africa is not shown because no significant signal
was detected there.

@@

pp <- CMAP() %>% 
  ReadNetCDF(subset = list(lat = c(-90, 0),
                           time = range(indices$time))) %>% 
  na.omit() %>% 
  normalise_coords() %>%
  .[, days := lubridate::days_in_month(time[1]), by = time]

@@

The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might
point to a link between the SAM and the Quasi Biennial Oscillation. There is
evidence of influence between the QBO and the Northern Annular Mode [e.g.
@holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would
be similarly affected. However, establishing this link would require further
research.
